The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-11-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Finished TODO's, article ready for reviews

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and RFI statistics are shown in the range 72--231~MHz and over various nights. RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV. After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DVB bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before reduction of observations. Advises are given to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Obervatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA. TODO mention something about the extent to which protections have been put in place for radio quietness at the MRO. Randall: ``I'm sure there's a document somewhere that describes it, otherwise you could contact Carol Wilson at CSIRO who's responsible for such things. (carol.wilson@csiro.au)''

% TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} (hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and conclude in \citet{offringa-rfi-distributions} that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citepalias{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citepalias{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Not all observations cover the DVB frequencies. (TODO this table excludes RF11 and RF12 DVB interference; needs fix)}\label{tbl:obs-list}
\begin{center}%
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
\hline
\multicolumn{4}{|r|}{\textbf{With uniform channel coverage}} & \textbf{1.13\%} & \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed noise in the high-resolution channels for each observation. This is the residual noise in the visibilities after RFI detection. A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from Digital Video Broadcasting (DVB), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DVB bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. }
\label{fig:dvb}
\end{figure*}

Because significant DVB interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DVB interference from the data statistics. An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DVB is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%. This is detailed in the last column of Table~\ref{tbl:obs-list}.

@@

\subsection{Distribution analysis}
In \citet{offringa-rfi-distributions}, using LOFAR it is shown that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are drawn in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that each observation covers different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines). The EoR high-band curves do not include the night that was contaminated by DVB RFI.

@@

\noindent\begin{figure*}%
% obs data: 1075641128
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-noflags}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-withflags}
\end{center}
\caption{DVB RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
\label{fig:dvb-burst}
\end{figure*}

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.

@@

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID setup. The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel. TODO mention samples/minute and that this is constant.

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DVB bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVB signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation. When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.

>>>>

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV). After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before reduction of observations. Some advise is deducted to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Obervatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

Several measures are in place to protect the radio quietness of the MRO: the frequency plan of the Australian Communications and Media Authority (ACMA) states that spectrum licenses cannot interfere within a radius of 70~km around the centre of the MRO. Within a radius of 150~km, consultation is required. Licenses for individual apparatus are controlled by the Radiocommunications Assignment and Licensing Instruction (RALI, document MS 32), which prescribes a 150~km restricted and 230~km consultation zone for frequencies 100--230~MHz, and therefore covers most of the bandwidth of the MWA. Cases that are not covered by the radio-quiet zone are: i) transmitters with frequencies below 70~MHz; ii) aircraft transmissions; iii) satellite transmissions; and iv) transmitters beyond 260~km from the centre of the MRO.

\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citeauthor{lofar-radio-environment} (\citeyear{lofar-radio-environment}; hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and conclude in \citet{offringa-rfi-distributions} that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial RFI detection at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citepalias{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citepalias{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). Occasionally, interference from digital TV (DTV) is visible. The DTV column displays the fraction of visibilities that were unusable because of the presence of DTV signals, and which the flagger does not flag. Not all observations cover the DTV frequencies.}\label{tbl:obs-list}
\begin{center}%
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DTV \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 1.29\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.14\%}\\
\hline
\multicolumn{4}{|r|}{\textbf{With uniform channel coverage}} & \textbf{1.13\%} & \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed noise in the high-resolution channels for each observation. This is the residual noise in the visibilities after RFI detection. A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, DTV interference has to be detected with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from digital TV (DTV), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DTV frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DTV bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. }
\label{fig:dvb}
\end{figure*}

Because significant DTV interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DTV transmitters are terrestrial, and the fact that only this night observes the DTV, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

Although the flagger does not adequately flag DTV, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DTV interference from the data statistics. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in the last column of Table~\ref{tbl:obs-list}. Because these frequencies correspond to the EoR redshift, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

\subsection{Distribution analysis}
In \citet{offringa-rfi-distributions}, using LOFAR it is shown that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are drawn in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that each observation covers different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines). The EoR high-band curves do not include the night that was contaminated by DTV RFI.

@@

\noindent\begin{figure*}%
% obs data: 1075641128
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-noflags}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-withflags}
\end{center}
\caption{DTV RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
\label{fig:dvb-burst}
\end{figure*}

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.

@@

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID setup. The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel. Assuming a 138~GFLOPS (Giga-floating-point operations per second) performance of the host computer, the RFI detection requires 25 FLOP/sample. When expressed as visibilities per time unit, the computational performance of the flagger is independent of the frequency resolution, time resolution and number of antennas. This number can therefore be extrapolated to other telescopes, although the performance of a pipeline which incorporates RFI detection will be strongly dependent on available IO performance and memory bandwidth.

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DTV signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation. When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.
